---
title: "Assignment 3"
author: "Aleksandar Maric, Akando Räderer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

library(dplyr)
library(ggplot2)
library(skimr)
library(lfe)
library(ivreg)
library(skimr)
library(car)
```

# Execise 1

**Estimating the simple linear regressions**

$$Y_{i,1} = \beta_0 + \beta_1 X_i + u_i$$ and $$Y_{i,2} = \alpha_0 + \alpha_1 X_i + v_i$$.

We define $Q_i = X_i - 10$.

### Proof that $\widehat{\delta_1} = \widehat{\beta_1}$

We estimate $$Y_{i,1} = \delta_0 + \delta_1 Q_i + u_i$$

We have $$Q_i = X_i - 10 \quad \Rightarrow \quad X_i = Q_i + 10.$$

Substituting into the original regression gives $$Y_{i,1} = \beta_0 + \beta_1 X_i + u_i = \beta_0 + \beta_1(Q_i + 10) + u_i.$$

Rearranging: $$Y_{i,1} = (\beta_0 + 10\beta_1) + \beta_1 Q_i + u_i.$$

Comparing this to $$Y_{i,1} = \delta_0 + \delta_1 Q_i + u_i,$$ we obtain $$\delta_0 = \beta_0 + 10\beta_1 \quad \text{and} \quad \delta_1 = \beta_1.$$

Hence, $$\widehat{\delta_1} = \widehat{\beta_1}.$$

Conclusion: adding or subtracting a constant from a regressor does not affect the OLS slope coefficient. It only changes the intercept. In this case, the relationship changes as follows:$$\widehat{\delta_0} = \widehat{\beta_0} + 10\widehat{\beta_1}.$$

# Exercise 2

## 1. Regression analysis: $medicine, travel\_time \quad \text{and} \quad  income\_parents$

### i) Import, exploration and cleaning

```{r}
df <- read.csv("distances_birthtown_up.csv")
skimr::skim(df)
```

After a first glance at the data set, I would suggest doing the same cleaning steps as in the exercise session (i.e. cleaning the unreported incomes and negative travel times):

```{r}
# Encode 'medicine' variable
df <- mutate(df, medicine = ifelse(medicine == "Yes", 1, 0))


# Set missing values for income variables
df <- mutate(df, income = ifelse(income == -9999999, NaN, income)) %>%
  mutate(income_parent1 = ifelse(income_parent1 == -9999999, NaN, income_parent1)) %>%
  mutate(income_parent2 = ifelse(income_parent2 == -9999999, NaN, income_parent2))


# Negative travel time
df <- mutate(df, travel_time = ifelse(travel_time < 0, NaN, travel_time))                    


# Omit missing variables 
df_sample <- df |>
  na.omit()
```

Now we generate the variable $income\_parents = income\_p1 + income\_p2$:

```{r}
df_sample <- df_sample |> mutate(income_parents = income_parent1 + income_parent2)
```

### ii) Regression Equation

To investigate the relationship of medicine with travel time and income parents we write the regression equation as follows:

$$medicine_i = \beta_0 + \beta_1 travel\_time_i + \beta_2 income\_parents_i + \epsilon_i$$

- $\beta_0$ is the predicted probability of working in medicine when travel time and parental income are both zero.

- $\beta_1$ measures the change in the probability of working in medicine associated with a one-unit increase in travel time, holding parental income constant.

- $\beta_2$ measures the change in the probability of working in medicine associated with a one-unit increase in parental income, holding travel time constant. 

$medicine_i$ is an indicator equal to 1 if individual $i$ works in medicine and 0 otherwise. $travel\_time_i$ measures travel time to Country B, and $income\_parents_i$ is the sum of both parents' incomes.

### iii) Estimation of Regression

We'll compare 

$$medicine_i = \beta_0 + \beta_1 travel\_time_i + \beta_2 income\_parents_i + \epsilon_i$$

to 

$$medicine_i = \delta_0 + \delta_1 income\_parents_i + v_i$$

```{r}
reg_joint <- lm(medicine ~ travel_time + income_parents, data = df_sample)
reg_short <- lm(medicine ~ income_parents, data = df_sample)

summary(reg_joint)
summary(reg_short)
```

In the joint regression from part (ii), the coefficient on income_parents is much smaller than in the regression from (iii). In the joint specification, the estimate is about 0.00020 and is not statistically significant at the 5% level, while in regression (iii) it is about 0.00081 and statistically significant. The coefficient on *travel_time* in the joint regression is negative and statistically significant, around -0.0498. Economically, this means that one additional hour of travel time is associated with about a 5 percentage point lower probability of working in medicine, holding parental income constant. A possible explanation is omitted variable bias in regression (iii): travel time is left out, although it is related to career choice and may also be correlated with parental income.



### iv) Better Explanation For The Research Question

Using *region* instead of *travel_time* would likely not help to answer the research question better. The main reason is that *travel_time* is a more direct measure of how easily an individual can access Country B and its medical industry. In contrast, *region* is only a broad categorical location measure and does not capture variation in actual distance or accessibility within regions. Two individuals in the same *region* may still face very different travel times (e.g. *region* = 11). This makes *travel_time* a more precise explanatory variable.

However, *travel_time* may not fully capture all real-world barriers to access, such as institutional or administrative constraints. Therefore, *region* could still capture broader local characteristics and may be useful as an additional control, but on its own it is less precise than *travel_time* for this research question.

### v) Correlation of *income_parents* with *travel_time*

**Hypotheses**

We estimate the following regression:

$$income\_parents_i = \gamma_0 + \gamma_1 travel\_time_i + u_i$$

To test whether *income_parents* is correlated with *travel_time*, we consider the following hypotheses:

$$H_0: \gamma_1 = 0$$

$$H_1: \gamma_1 \neq 0$$

The null hypothesis states that parental income is not correlated with travel time, while the alternative states that there is a correlation.

$$t=\frac{\hat{\gamma}_1}{se(\hat{\gamma}_1)}$$

```{r}
df_test <- df_sample

fit_corr <- felm(income_parents ~ travel_time, data = df_test)
summary(fit_corr)

cor.test(df_test$income_parents, df_test$travel_time)
```

The estimated coefficient on *travel_time* is about -17.11, with a standard error of 0.2937, giving a t-statistic of -58.27 and a p-value below 0.01. Hence, we reject $H_0$ at the 5% level. *We could also compare it to the critical value $t_{crit} = -1.96$ from `qt(0.025, 12804)` and see immediately that we're far away from it.*

**Conclusion**

There is strong statistical evidence that *income_parents* and *travel_time* are correlated, and the relationship is negative. Individuals with higher parental income tend to live closer to Country B. Omitting *travel_time* leads to omitted variable bias in part (iii). The estimated effect of parental income alone captures both its direct effect and part of the effect of travel time.

### vi) Joint Significance of *travel_time* and *income_parents*

To test whether they are jointly significant we test 
 
$$H_0:\beta_1=0,\ \beta_2=0$$ against 
$$H_A:\text{at least one of the two coefficients is different from 0.}$$
```{r}
fit_medicine <- felm(medicine ~ travel_time + income_parents, data = df_sample)
summary(fit_medicine)

linearHypothesis(fit_medicine, c("travel_time = 0", "income_parents = 0"))
```
Or by hand from $$F=\frac{(SSR_r-SSR_{ur})/q}{SSR_{ur}/(n-k-1)}$$

```{r}
fit_ur <- lm(medicine ~ travel_time + income_parents, data = df_sample) # unrestricted model containing both coefficients
fit_r  <- lm(medicine ~ 1, data = df_sample) # model only containing the intercept --> q = 2 (restrictions)

ssr_r  <- sum(fit_r$residuals^2)
ssr_ur <- sum(fit_ur$residuals^2)

q <- 2
k <- 2
n <- nobs(fit_ur)

Fstat <- ((ssr_r - ssr_ur)/q) / (ssr_ur/(n-k-1))
Fstat
qf(0.95, q, n-k-1)
```

The F-statistic is larger than the critical value at the 95th percentile for a test at the 5% significance level. *income_parents* and *travel_time* are jointly significant. This means that the two variables together help explain the probability of working in medicine.