---
title: "Assignment_5_Maric_Räderer"
author: "Akando Räderer - 23-615-669, Aleksandar Maric - 24-616-427"
output:
  html_document: default
  pdf_document: default
format:  
  html:
    toc: true                # Adds a Table of Contents on the side
    toc-depth: 4             # Includes headings up to level 4 (####) in the TOC
    number-sections: false   # Numbers your sections automatically if set to true
    code-fold: false         # Keeps your code visible (set to true to hide it)
    code-annotations: hover  # Reveals code annotations when hovering
    theme: cosmo             # A clean, built-in look
    embed-resources: true    # Packages everything into a single self-contained HTML file
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1: Descriptive Analysis

```{r}
# download and clean data
library(tidyverse)
library(skimr)

data <- read_csv("education_income-3.csv")

data_clean <- data %>%
  filter(
    !is.na(id),
    !duplicated(id),
    year %in% c(2000, 2010),
    college %in% c(0, 1),
    income >= 0,
    scholarship >= 0,
    hhincome >= 0,
    race %in% c(0, 1)
  ) %>%
  mutate(
    white = race,
    low_income = ifelse(hhincome < median(hhincome, na.rm = TRUE), 1, 0)
  )

skim(data_clean)


data_clean %>%
  summarise(
    observations = n(),
    duplicate_ids = sum(duplicated(id)),
    min_year = min(year),
    max_year = max(year),
    invalid_college = sum(!college %in% c(0, 1)),
    min_income = min(income),
    min_scholarship = min(scholarship),
    min_hhincome = min(hhincome),
    invalid_race = sum(!race %in% c(0, 1)),
    missing_values = sum(is.na(.))
  )

```

```{r}
# calculate income means
income_college <- data_clean %>%
    filter(college == 1) %>%
    pull(income)

income_non_college <- data_clean %>%
    filter(college == 0) %>%
    pull(income)



mean_all <- mean(data_clean$income)

mean_college <- mean(income_college)

mean_non_college <- mean (income_non_college)

print(mean_all)
print(mean_college)
print(mean_non_college)


mean_difference <- mean_college - mean_non_college

print(mean_difference)
```

-\> explain ...

# Question 2: Assignment Mechanism

## (a) 

Let $D_i$ denote the treatment indicator, where

$$
D_i =
\begin{cases}
1 & \text{if individual } i \text{ attends college},\\
0 & \text{otherwise}.
\end{cases}
$$

Let $Y_{1i}$ denote the potential income of individual $i$ if they attend college, and let $Y_{0i}$ denote the potential income if they do not attend college. The observed outcome is therefore

$$
Y_i = D_iY_{1i} + (1-D_i)Y_{0i}.
$$

The observed raw mean difference is

$$
E(Y_i \mid D_i = 1) - E(Y_i \mid D_i = 0).
$$

Using potential outcomes, this can be written as

$$
E(Y_{1i} \mid D_i = 1) - E(Y_{0i} \mid D_i = 0).
$$

Adding and subtracting $E(Y_{0i} \mid D_i = 1)$ gives

$$
E(Y_{1i} \mid D_i = 1) - E(Y_{0i} \mid D_i = 0)
$$

$$
=
\underbrace{E(Y_{1i} - Y_{0i} \mid D_i = 1)}_{\text{ATET}}
+
\underbrace{E(Y_{0i} \mid D_i = 1) - E(Y_{0i} \mid D_i = 0)}_{\text{Selection bias}}.
$$

The first term is the average treatment effect on the treated. The second term captures whether college attendees and non-attendees would have had different incomes even in the absence of college attendance.

## (b)

$\hat{\beta}_1$ does not necessarily measures the causal effect of college attendance on income. The reason is that college attendance is unlikely to be randomly assigned. Individuals who attend college may differ systematically from non-attendees in characteristics that also affect future income, such as household income, ability, motivation, or family background. Therefore, the error term $u_i$ may contain omitted determinants of income that are correlated with $college_i$. Formally, this means

$$
E(u_i \mid college_i) \neq 0.
$$

As a result, the OLS estimate $\hat{\beta}_1$ is likely biased and captures both the causal effect of college attendance and selection bias.

## (c)

A key observed driver of selection into college is household income, $hhincome_i$. Students from higher-income households may be more likely to attend college because they face fewer financial constraints and may have better access to educational resources. At the same time, household income may also be correlated with future income through family background, networks, school quality, or other socioeconomic advantages. Therefore, $hhincome_i$ can affect both college attendance and later income, making it a plausible source of selection bias.

## (d)

```{r}
# Regression without controls
ols_simple <- lm(income ~ college, data = data_clean)

# Regression with observed controls
ols_controls <- lm(income ~ college + hhincome + race + factor(year),
                   data = data_clean)

summary(ols_simple)
summary(ols_controls)

# Compare the estimated coefficient on college
coef_comparison <- tibble(
  Model = c("Without controls", "With controls"),
  College_coefficient = c(
    coef(ols_simple)["college"],
    coef(ols_controls)["college"]
  )
)

coef_comparison
```

The estimated coefficient on $college_i$ falls once household income, race, and year are added as controls. In the simple regression, the coefficient is approximately $41.51$, while in the controlled regression it decreases to approximately $28.39$. This suggests that part of the raw difference in income was due to observable differences between college and non-college individuals, rather than college attendance itself. However, adding these controls only addresses selection on observables. The estimate can still be biased if unobserved factors, such as ability, motivation, parental education, or school quality, affect both college attendance and later income. Therefore, the controlled OLS estimate is more credible than the raw comparison, but it still does not necessarily identify the causal effect of college attendance.

## (e)

If scholarship amounts are randomly assigned, then scholarship variation is independent of students' potential outcomes and pre-existing characteristics. Therefore, students receiving higher and lower scholarships should be comparable on average in terms of ability, motivation, family background, and other determinants of future income.

In terms of the decomposition from part (a), random assignment implies that the counterfactual income without college is the same across treatment groups:

$$
E(Y_{0i} \mid D_i = 1) = E(Y_{0i} \mid D_i = 0).
$$

Therefore, the selection bias term,

$$
E(Y_{0i} \mid D_i = 1) - E(Y_{0i} \mid D_i = 0),
$$

equals zero. The remaining observed difference can then be interpreted causally, for the variation in college attendance induced by the randomized scholarship.

## (f)

```{r}
library(lfe)

# OLS with controls for comparison
ols_controls <- felm(income ~ college + hhincome + race + factor(year),
                     data = data_clean)

summary(ols_controls, robust = TRUE)

# First stage: effect of scholarship on college attendance
first_stage <- felm(college ~ scholarship + hhincome + race + factor(year),
                    data = data_clean)

summary(first_stage, robust = TRUE)

# IV / 2SLS using scholarship as an instrument for college
iv_model <- felm(income ~ hhincome + race + factor(year) |
                   0 |
                   (college ~ scholarship),
                 data = data_clean)

summary(iv_model, robust = TRUE)

coef(ols_controls)
coef(iv_model)
names(coef(iv_model))

iv_college_name <- names(coef(iv_model))[grepl("college", names(coef(iv_model)))]
iv_college_coef <- coef(iv_model)[iv_college_name]

coef_comparison <- tibble(
  Model = c("OLS with controls", "IV / 2SLS"),
  College_coefficient = c(
    coef(ols_controls)["college"],
    iv_college_coef
  )
)

coef_comparison

```

In the first stage, $scholarship_i$ is positively and significantly related to college attendance. The coefficient is approximately $0.0098$, meaning that an additional 1,000 USD in scholarship support increases the probability of attending college by about 0.98 percentage points, conditional on household income, race, and year. The first-stage $F$-statistic is well above 10, so the instrument is relevant.

The IV estimate of the effect of college on income is approximately $21.08$, compared to $28.39$ in the controlled OLS regression. Since scholarship amounts are randomly assigned, the independence assumption is more credible. However, a causal interpretation still requires the exclusion restriction: scholarships must affect later income only through college attendance. Under this assumption, the IV estimate can be interpreted as a causal LATE for compliers.

# Question 3: Effect Heterogeneity

## (a) 

The regression equation is

$$
income_i = \beta_0 + \beta_1 college_i + \beta_2 white_i
+ \beta_3(college_i \times white_i) + u_i,
$$

where $white_i = 1$ for white individuals and $white_i = 0$ otherwise.

Here, $\beta_1$ measures the college income premium for non-white individuals. The coefficient $\beta_3$ captures treatment effect heterogeneity by race. It measures how much the college income premium differs for white individuals relative to non-white individuals. Therefore, the marginal effect of college is

$$
\beta_1
$$

for non-white individuals, and

$$
\beta_1 + \beta_3
$$

for white individuals. If $\beta_3 = 0$, there is no difference in the college income premium across race groups.

## (b)

```{r}
# Make sure the race group is coded as white = 1 and white = 0 otherwise
data_clean <- data_clean %>%
  mutate(white = race)

# Estimate the interaction model from Question 3(a)
heterogeneity_model <- lm(income ~ college * white, data = data_clean)

# Show full regression output
summary(heterogeneity_model)

# Show a cleaner coefficient table
coef_table_3b <- summary(heterogeneity_model)$coefficients

round(coef_table_3b, 3)
```

The regression from part (a) is estimated by OLS. The interaction term $college_i \times white_i$ is included to allow the return to college to differ between white and non-white individuals.

## (c)

```{r}
# Extract coefficients
beta_1 <- coef(heterogeneity_model)["college"]
beta_3 <- coef(heterogeneity_model)["college:white"]

# Marginal effects of college by race group
effect_nonwhite <- beta_1
effect_white <- beta_1 + beta_3

effects_by_race <- tibble(
  Group = c("white != 1", "white == 1"),
  Marginal_effect_college = c(effect_nonwhite, effect_white)
)
 print(effects_by_race)
```

The estimated marginal effect of college attendance is approximately $38.03$ for individuals with $white_i \neq 1$. For individuals with $white_i = 1$, the estimated marginal effect is

$$
\hat{\beta}_1 + \hat{\beta}_3 = 38.03 + 1.17 = 39.20.
$$

Since income is measured in 1,000 USD, this means that college attendance is associated with approximately 38,030 USD higher income for non-white individuals and approximately 39,200 USD higher income for white individuals.

## (d)

```{r}
# Extract interaction coefficient and p-value
interaction_result <- summary(heterogeneity_model)$coefficients["college:white", ]

interaction_result

# Cleaner output
tibble(
  Coefficient = "college:white",
  Estimate = interaction_result["Estimate"],
  Std_Error = interaction_result["Std. Error"],
  t_value = interaction_result["t value"],
  p_value = interaction_result["Pr(>|t|)"]
)
```

The difference in the college income premium across race groups is captured by the interaction coefficient $\beta_3$ on $college_i \times white_i$. The estimated interaction coefficient is approximately $1.17$, with a p-value of approximately $0.318$. Therefore, we do not reject

$$
H_0: \beta_3 = 0.
$$

This means that the estimated college premium is slightly higher for white individuals, but the difference is not statistically significant. Hence, the data do not provide strong evidence that the return to college differs across race groups.

# Question 4: Probit and Logit Models

## (a) 

A limitation of using OLS when the dependent variable is binary is that the fitted values are interpreted as predicted probabilities, but OLS does not restrict them to lie between 0 and 1. Therefore, the model may predict probabilities below 0 or above 1, which has no meaningful probability interpretation. This is why logit and probit models are often preferred for binary outcomes.

## (b)

```{r}
# Probit model
probit_model <- glm(
  college ~ hhincome + race,
  data = data_clean,
  family = binomial(link = "probit")
)

summary(probit_model)

# Logit model
logit_model <- glm(
  college ~ hhincome + race,
  data = data_clean,
  family = binomial(link = "logit")
)

summary(logit_model)
```

The probit and logit models give the same qualitative result. In both models, the coefficient on $hhincome_i$ is positive and statistically significant. This suggests that individuals from higher-income households are more likely to attend college.

The coefficient on $race_i$ is negative and statistically significant in both models. This means that individuals with $race_i = 1$ have a lower predicted probability of attending college, conditional on household income.

The logit coefficients are larger in absolute value than the probit coefficients because the two models use different link functions and are measured on different scales. Therefore, the raw coefficient sizes should not be interpreted directly as changes in probability. However, both models agree on the direction of the effects.

# Question 5: Marginal Effects

## (a) 

## (b) 

## (c)
