---
title: 'Data Analytics II: Assignment 5'
author: "Aleksandar Maric, Akando Räderer"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(skimr)
library(lfe)
```

# Data Cleaning

```{r}
# Load the data
df <- read.csv("education_income-3.csv")

# Summarize the data
skim(df)
```

```{r}
# Set NA for college column, where it's neither 1 or 0
df_clean <- df |>
  mutate(
    college = ifelse(college == -1, NA, college),
    income = ifelse(income == -9999999, NA, income),
    hhincome = ifelse(hhincome == -9999999, NA, hhincome),
   white = ifelse(race == 1, 1, 0), # white == 1, non-white otherwise, for Q3
    low_income = ifelse(
      is.na(hhincome),
      NA,
      ifelse(hhincome < median(hhincome, na.rm = TRUE), 1, 0)
    )
  ) |>
  filter(
    !is.na(college),
    !is.na(income),
    !is.na(hhincome)
  )

skim(df_clean)
```

# Question 1: Descriptive Analysis

```{r}
# Mean incomes
mean_income <- df_clean |>
  group_by(college) |>
  summarise(
    n = n(),
    mean_income = mean(income, na.rm = TRUE)
  )

mean_income

mean_difference <- mean_income |>
  summarise(
    raw_difference = mean_income[college == 1] - mean_income[college == 0]
  )

mean_difference
```

The raw mean income difference is about 41.51, meaning that college attendees earn on average about 41.51 thousand USD more than non-attendees. However, this difference should not be interpreted as the causal effect of attending college. College attendance is likely not randomly assigned: individuals who attend college may differ systematically from non-attendees in household income, scholarship support, race, ability, motivation, family background, or other factors. These factors may affect both college attendance and later income, creating **selection bias**. Therefore, the raw mean difference is only a descriptive association, not a causal estimate.

# Question 2: Assignment Mechanism

## a) Potential outcomes, observed mean difference, ATET and selection bias

Let $D_i$ denote the treatment indicator, where

$$
D_i =
\begin{cases}
1 & \text{if individual } i \text{ attends college},\\
0 & \text{otherwise}.
\end{cases}
$$

Let $Y_{1i}$ denote the potential income of individual $i$ if they attend college, and let $Y_{0i}$ denote the potential income if they do not attend college. The observed outcome is therefore

$$
Y_i = D_iY_{1i} + (1-D_i)Y_{0i}.
$$

The observed raw mean difference is

$$
E(Y_i \mid D_i = 1) - E(Y_i \mid D_i = 0).
$$

Using potential outcomes, this equals

$$
E(Y_{1i} \mid D_i = 1) - E(Y_{0i} \mid D_i = 0),
$$

because for treated individuals we observe $Y_{1i}$ and for untreated individuals we observe $Y_{0i}$.

To decompose this expression, we add and subtract the same term, $E(Y_{0i} \mid D_i = 1)$:

$$
E(Y_{1i} \mid D_i = 1) - E(Y_{0i} \mid D_i = 0)
$$

$$
=
E(Y_{1i} \mid D_i = 1)
-
E(Y_{0i} \mid D_i = 1)
+
E(Y_{0i} \mid D_i = 1)
-
E(Y_{0i} \mid D_i = 0).
$$

Therefore,

$$
E(Y_i \mid D_i = 1) - E(Y_i \mid D_i = 0)
$$

$$
=
\underbrace{E(Y_{1i} - Y_{0i} \mid D_i = 1)}_{\text{ATET}}
+
\underbrace{E(Y_{0i} \mid D_i = 1) - E(Y_{0i} \mid D_i = 0)}_{\text{Selection bias}}.
$$

The first term is the average treatment effect on the treated (ATET), i.e. the average causal effect of college attendance for people who actually attended college ($D_i = 1$). The second term captures whether college attendees and non-attendees would have had different incomes even if nobody had attended college (selection bias).

## b) $\hat \beta_1$ interpretation

$\hat{\beta}_1$ does not necessarily measure the causal effect of college attendance on income. The reason is that college attendance is unlikely to be randomly assigned. Individuals who attend college may differ systematically from non-attendees in characteristics that also affect future income, such as household income, ability, motivation, or family background. Therefore, the error term $u_i$ may contain omitted determinants of income that are correlated with $college_i$. Formally, this means

$$
E(u_i \mid college_i) \neq 0.
$$

As a result, the OLS estimate $\hat{\beta}_1$ is likely biased and captures both the causal effect of college attendance and selection bias.

## c) Key driver of selection

A key observed driver of selection into college is household income, $hhincome_i$. Students from higher-income households may be more likely to attend college because they face fewer financial constraints and may have better access to educational resources. At the same time, household income may also be correlated with future income through family background, networks, school quality, or other socioeconomic advantages. Therefore, $hhincome_i$ can affect both college attendance and later income, making it a plausible source of selection bias.

## d) Estimation of regression

```{r}
# Regression without controls
ols_simple <- lm(income ~ college, data = df_clean)

# Regression with observed controls
ols_controls <- lm(income ~ college + hhincome + race + factor(year),
                   data = df_clean)

summary(ols_simple)
summary(ols_controls)

# Compare the estimated coefficient on college
coef_comparison <- tibble(
  Model = c("Without controls", "With controls"),
  College_coefficient = c(
    coef(ols_simple)["college"],
    coef(ols_controls)["college"]
  )
)

coef_comparison
```

The estimated coefficient on $college_i$ falls once household income, race, and year are added as controls. In the simple regression, the coefficient is approximately $41.51$, while in the controlled regression it decreases to approximately $28.39$. This suggests that part of the raw difference in income was due to observable differences between college and non-college individuals, rather than college attendance itself. However, adding these controls only addresses selection on observables. The estimate can still be biased if unobserved factors, such as ability, motivation, parental education, or school quality, affect both college attendance and later income. Therefore, the controlled OLS estimate is more credible than the raw comparison, but it still does not necessarily identify the causal effect of college attendance.

## e) Randomly assigned scholarships

If scholarship amounts are randomly assigned, scholarship is independent of students’ potential outcomes and pre-treatment characteristics. Therefore, students who receive larger or smaller scholarships should not differ systematically in unobserved ability, motivation, family background, or other determinants of future income.

This does not mean that college attendance itself is randomly assigned, because students still decide whether to attend college. Therefore, random scholarships do not directly imply

$$
E(Y_{0i} \mid D_i = 1) = E(Y_{0i} \mid D_i = 0).
$$

However, random scholarship assignment creates exogenous variation in college attendance. If scholarships affect the probability of attending college, the relevance assumption is satisfied. Because the scholarship amount is randomly assigned, the exogeneity assumption is more plausible. Under these IV assumptions, scholarship can be used as an instrument to estimate the causal effect of college attendance on income.

## f) 2SLS Estimation model

```{r}
# OLS with controls for comparison
ols_controls <- felm(income ~ college + hhincome + race + factor(year),
                     data = df_clean)

summary(ols_controls, robust = TRUE)

# First stage: effect of scholarship on college attendance
first_stage <- felm(college ~ scholarship + hhincome + race + factor(year),
                    data = df_clean)

summary(first_stage, robust = TRUE)

# IV / 2SLS using scholarship as an instrument for college
iv_model <- felm(income ~ hhincome + race + factor(year) |
                   0 |
                   (college ~ scholarship),
                 data = df_clean)

summary(iv_model, robust = TRUE)

coef(ols_controls)
coef(iv_model)
names(coef(iv_model))

iv_college_name <- names(coef(iv_model))[grepl("college", names(coef(iv_model)))]
iv_college_coef <- coef(iv_model)[iv_college_name]

coef_comparison <- tibble(
  Model = c("OLS with controls", "IV / 2SLS"),
  College_coefficient = c(
    coef(ols_controls)["college"],
    iv_college_coef
  )
)

coef_comparison

```

In the first stage, $scholarship_i$ is positively and significantly related to college attendance. The coefficient is approximately $0.0098$, so an additional 1,000 USD in scholarship support increases the probability of attending college by about 0.98 percentage points, conditional on household income, race, and year. This supports relevance: $\operatorname{Cov}(scholarship, college) \neq 0$.

The IV estimate of the effect of college on income is approximately $21.08$, compared to approximately $28.39$ in the controlled OLS regression. The smaller IV estimate suggests that controlled OLS may still contain upward selection bias. A causal interpretation additionally requires exogeneity: scholarship must not be correlated with unobserved determinants of income and must affect income only through college attendance.

# Question 3: Effect Heterogeneity

## a) Regression 

The regression equation is

$$
income_i = \beta_0 + \beta_1 college_i + \beta_2 white_i
+ \beta_3(college_i \times white_i) + u_i,
$$

where $white_i = 1$ for white individuals and $white_i = 0$ otherwise.

Here, $\beta_1$ measures the college income premium for non-white individuals. The coefficient $\beta_3$ captures treatment effect heterogeneity by race. It measures how much the college income premium differs for white individuals relative to non-white individuals. Therefore, the marginal effect of college is

$$
\frac{\partial income_i}{\partial college_i} = \beta_1
$$

for non-white individuals, and

$$
\frac{\partial income_i}{\partial college_i} =\beta_1 + \beta_3
$$

for white individuals. If $\beta_3 = 0$, there is no difference in the college income premium across race groups.

## b) Estimation 

```{r}
# Estimation
heterogeneity_model <- lm(income ~ college * white, data = df_clean)

# Full regression output
summary(heterogeneity_model)

# A cleaner coefficient table
coef_table_3b <- summary(heterogeneity_model)$coefficients

round(coef_table_3b, 3)
```

The regression from part (a) is estimated by OLS. The interaction term $college_i \times white_i$ is included to allow the return to college to differ between white and non-white individuals.

## c) Marginal effect of college on income

```{r}
# Extract coefficients
beta_1 <- coef(heterogeneity_model)["college"]
beta_3 <- coef(heterogeneity_model)["college:white"]

# Marginal effects of college by race group
effect_nonwhite <- beta_1
effect_white <- beta_1 + beta_3

effects_by_race <- tibble(
  Group = c("white != 1", "white == 1"),
  Marginal_effect_college = c(effect_nonwhite, effect_white)
)
 print(effects_by_race)
```

The estimated marginal effect of college attendance is approximately $38.03$ for individuals with $white_i \neq 1$. For individuals with $white_i = 1$, the estimated marginal effect is

$$
\hat{\beta}_1 + \hat{\beta}_3 = 38.03 + 1.17 = 39.20.
$$

Since income is measured in 1,000 USD, this means that college attendance is associated with approximately 38,030 USD higher income for non-white individuals and approximately 39,200 USD higher income for white individuals.

## d) Significance

```{r}
# Extract interaction coefficient and p-value
interaction_result <- summary(heterogeneity_model)$coefficients["college:white", ]

interaction_result
```

The difference in the college income premium across race groups is captured by the interaction coefficient $\beta_3$ on $college_i \times white_i$. The estimated interaction coefficient is approximately $1.17$, with a p-value of about $0.318$. Therefore, we do not reject

$$
H_0: \beta_3 = 0.
$$

This means that the estimated college premium is slightly higher for white individuals, but the difference is not statistically significant. Hence, the data do not provide strong evidence that the return to college differs across race groups.

# Question 4: Probit and Logit Models

## a) Limitation of using OLS with binary dependent variable

A limitation of using OLS when the dependent variable is binary is that the fitted values are interpreted as predicted probabilities, but OLS does not restrict them to lie between 0 and 1. Therefore, the model may predict probabilities below 0 or above 1, which has no meaningful probability interpretation. This is why Logit and Probit models are often preferred for binary outcomes.

## b) Estimation of Logit / Probit Model

```{r}
# Probit model
probit_model <- glm(
  college ~ hhincome + race,
  data = df_clean,
  family = binomial(link = "probit")
)

summary(probit_model)

# Logit model
logit_model <- glm(
  college ~ hhincome + race,
  data = df_clean,
  family = binomial(link = "logit")
)

summary(logit_model)
```

The Probit and Logit models give the same qualitative result. In both models, the coefficient on $hhincome_i$ is positive and statistically significant. This suggests that individuals from higher-income households are more likely to attend college.

The coefficient on $race_i$ is negative and statistically significant in both models. This means that individuals with $race_i = 1$ have a lower predicted probability of attending college, conditional on household income.

The Logit coefficients are larger in absolute value than the Probit coefficients because the two models use different link functions and are measured on different scales. Therefore, the raw coefficient sizes should not be interpreted directly as changes in probability. However, both models agree on the direction of the effects.

# Question 5: Marginal Effects

## a) APE of a continuous $x_j$ in a Probit model

In the probit model,

$$
P(college_i = 1 \mid x_i) = \Phi(x_i'\beta),
$$

where $\Phi(\cdot)$ is the standard normal CDF. For a continuous variable $x_j$, the partial effect for individual $i$ is

$$
\frac{\partial P(college_i = 1 \mid x_i)}{\partial x_{ij}}
=
\phi(x_i'\beta)\beta_j,
$$

where $\phi(\cdot)$ is the standard normal density. The Average Partial Effect is therefore

$$
APE_j =
\frac{1}{n}
\sum_{i=1}^n
\phi(x_i'\hat{\beta})\hat{\beta}_j.
$$

## b) Computation and interpretation

```{r}
probit_model <- glm(
  college ~ hhincome + white, # using 'white' instead of 'race' according to assignment
  data = df_clean,
  family = binomial(link = "probit")
)

summary(probit_model)

# Predicted linear index: beta_0 + beta_1*hhincome + beta_2*race
xb <- predict(probit_model, type = "link")

# Extract coefficient on hhincome
beta_hhincome <- coef(probit_model)["hhincome"]

# APE formula for probit:
# APE_j = (1/n) * sum_i phi(x_i'beta) * beta_j
ape_hhincome <- mean(dnorm(xb) * beta_hhincome)

ape_hhincome
```

The estimated Average Partial Effect of $hhincome$ is approximately

$$
0.00318.
$$

Thus, an increase in household income by 1,000 USD is associated with an increase in the probability of attending college by about 0.318 percentage points on average, holding race constant. This is an average effect because, in a probit model, the marginal effect depends on the values of the covariates.

## c) Predicted probability

```{r}
individual <- data.frame(
  hhincome = 80,
  white = 1
)

predicted_probability <- predict(
  probit_model,
  newdata = individual,
  type = "response"
)

predicted_probability
```


For an individual with

$$
hhincome = 80
$$

and

$$
white = 1,
$$

the predicted probability of attending college is approximately

$$
0.740.
$$

Therefore, the model predicts a college attendance probability of about 74.0%. This value lies between 0 and 1, as expected in a probit model.


