---
title: "Assignment 3"
author: "Aleksandar Maric, Akando Räderer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

library(dplyr)
library(ggplot2)
library(skimr)
library(lfe)
library(ivreg)
library(skimr)
library(car)
```

# Execise 1

Estimating the simple linear regressions

$$Y_{i,1} = \beta_0 + \beta_1 X_i + u_i$$ and $$Y_{i,2} = \alpha_0 + \alpha_1 X_i + v_i$$.

We define $Q_i = X_i - 10$.

## Proof that $\widehat{\delta_1} = \widehat{\beta_1}$

We estimate $$Y_{i,1} = \delta_0 + \delta_1 Q_i + u_i$$

We have $$Q_i = X_i - 10 \quad \Rightarrow \quad X_i = Q_i + 10.$$

Substituting into the original regression gives $$Y_{i,1} = \beta_0 + \beta_1 X_i + u_i = \beta_0 + \beta_1(Q_i + 10) + u_i.$$

Rearranging: $$Y_{i,1} = (\beta_0 + 10\beta_1) + \beta_1 Q_i + u_i.$$

Comparing this to $$Y_{i,1} = \delta_0 + \delta_1 Q_i + u_i,$$ we obtain $$\delta_0 = \beta_0 + 10\beta_1 \quad \text{and} \quad \delta_1 = \beta_1.$$

Hence, $$\widehat{\delta_1} = \widehat{\beta_1}.$$

Conclusion: adding or subtracting a constant from a regressor does not affect the OLS slope coefficient. It only changes the intercept. In this case, $$\widehat{\delta_0} = \widehat{\beta_0} + 10\widehat{\beta_1}.$$

# Exercise 2

## 1. Regression analysis: $medicine, travel\_time and income\_parents$

### i Import, exploration and cleaning

```{r}
df <- read.csv("distances_birthtown_up.csv")
skimr::skim(df)
```
After a first glance at the data set, I would suggest doing the same cleaning steps as in the exercise session (i.e. cleaning the unreported incomes and negative travel times):

```{r}
# Encode 'medicine' variable
df <- mutate(df, medicine = ifelse(medicine == "Yes", 1, 0))


# Set missing values for income variables
df <- mutate(df, income = ifelse(income == -9999999, NaN, income)) %>%
  mutate(income_parent1 = ifelse(income_parent1 == -9999999, NaN, income_parent1)) %>%
  mutate(income_parent2 = ifelse(income_parent2 == -9999999, NaN, income_parent2))


# Negative travel time
df <- mutate(df, travel_time = ifelse(travel_time < 0, NaN, travel_time))                    


# Omit missing variables 
df_sample <- df |>
  na.omit()
```













