---
title: "Assignment 1"
author: "Aleksandar Maric, Akando Räderer, Daniel Margolin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(ggplot2)
```

## Exercise 1

### Cleaning the dataset

First we import the dataset in R and take a glimpse at the `summary()`.

```{r}
data <- read.csv("Assignment1.csv")
summary(data)
```

```{r}
skimr::skim(data)
```

------------------------------------------------------------------------

We know that the dataset contains information (5190 entries) on doctor visits in Johto with several variables. Now we handle extreme values.

At first sight we have 519 missing values for freepoor which we will remove with the assumption of them MCAR.

```{r}
data <- data |> filter(!is.na(freepoor))
```

After a closer look, we find that there are several entries for the `reduced` variable with $99$ which makes no sense if the question regarding this variable asks for days of reduced activity in the past two weeks. We can also observe one `age_squared` that is quite extreme ($82810000 \rightarrow \sqrt{82810000} = 9100$). This could of course be a typo for age 91, but it's safer to remove the entry.

In the following chunk, we change the "none" entries to "$0$" and the "entire period" to "$14$", then convert to numeric and filter for $\leq 14$. Finally, we explicitly filter out impossible values for `age_squared` by bounding the maximum realistic age (e.g., 110 years, meaning `age_square` $\leq 12100$).

```{r}
data <- data |> 
  mutate(
    reduced = case_when(
      reduced == "none" ~ "0",
      reduced == "entire period" ~ "14",
      TRUE ~ reduced  
    ),
    reduced = as.numeric(reduced)
  ) |>
  filter(reduced <= 14) |>
  filter(age_squared <= 12100)
```

------------------------------------------------------------------------

### Monthly Income

To compute the average monthly income and its variance, we should first use the correct units. According to the assignment, it's the annual income in tens of thousands of dollars. We therefore just multiply the whole column by 10'000 and divide by 12 (and create a new column):

```{r}
data <- data |> mutate(monthly_income = (income * 10000) / 12)
```

We then use `mean()` $\frac{1}{n}\sum_{i=1}^n X_i$ for the average monthly income and `var()` $Var(X) = \mathbb E [(X-\mathbb E[X])^2]$ for the variance (and remember that the standard deviation (`sd() = sqrt(var())`) is just its root $\sigma_X = \sqrt{Var(X)}$).

```{r}
mean(data$monthly_income, na.rm=TRUE)

var(data$monthly_income, na.rm=TRUE)

sd(data$monthly_income, na.rm=TRUE)
#sqrt(variance_monthly_income)
```

------------------------------------------------------------------------

#### Correlation Income vs. Health

***Ex-ante:*** We assume that a higher monthly income is associated with better overall health. However, since we're working with a GHQ score using Goldberg's method (where a higher score = worse health), we expect to see a negative statistical correlation between the two variables.

```{r}
cor(data$monthly_income, data$health)
```

***Ex-post:*** The correlation coefficient of -0.08 indicates a statistically significant but weak negative relationship between monthly income and the GHQ health score.

<br>

## Exercise 2

### Creating relevant datasets

```{r, warning=FALSE}
data_private <- data |> 
  select(health, monthly_income, private) |> 
  mutate(
    insurance_type = factor(
      ifelse(private == 1, "private", "non-private"), 
      levels = c("non-private", "private")
    )
  )

data_age <- data |> 
  select(health, monthly_income, age_squared) |> 
  mutate(
    age_class = factor(
      ifelse(age_squared > median(age_squared), "Above Median", "Below Median"),  
      levels = c("Below Median", "Above Median")
    )
  )
```

### Generating the graphs

```{r}
ggplot(data_private, aes(x=health, y=monthly_income)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  facet_wrap(~ insurance_type) + 
  labs(
    title = "Correlation between Health and Income by Insurance Type",
    x = "Health",
    y = "Income"
  ) + 
  theme_minimal()

# For the visual correlation
cor_by_group1 <- data_private |>
  group_by(insurance_type) |>
  summarise(correlation = cor(monthly_income, health, use = "complete.obs"))

print(cor_by_group1)
```

```{r}
ggplot(data_age, aes(x=health, y=monthly_income)) + 
  geom_point() +
  geom_smooth(method = "lm") + 
  facet_wrap(~ age_class) + 
  labs(
    title = "Correlation between Health and Income by Age Class",
    x = "Health",
    y = "Income"
  ) + 
  theme_minimal()

cor_by_group2 <- data_age |>
  group_by(age_class) |>
  summarise(correlation = cor(monthly_income, health, use = "complete.obs"))

print(cor_by_group2)
```

```{r}
ggplot(data_private, aes(x=health)) + 
  geom_histogram() + 
  facet_wrap(~ insurance_type) + 
  labs(
    title = "Histogram of Health by Insurance Type",
    x = "Health",
    y = "Count"
  ) + 
  theme_minimal()
```

### Conclusion

The correlation of these graphs are all small and negative ($\sim-0.08, \sim-0.06, \sim-0.08, \sim-0.08$). In a simple bivariate regression $r^2=R^2$. This implies that income explains less than $1\%$ of the variation in health. This suggests that income alone has very limited explanatory power for health in this sample.

In the case of the insurance types, the correlation is slightly negative for both types and the slopes are similar. The individuals with private insurance appear to have slightly better average levels of health.

The graphs by age show similar results. The negative association between income and health persists across age groups. Individuals below the median age appear to have better overall health than those above the median.

The histograms look similar and follow the same trend. The private insurance group is smaller overall, suggesting fewer privately insured individuals in the sample. The similar shapes indicate only minor differences in the distribution of health. We can draw the following conclusion:

-   Income shows only a minimal correlation with health, suggesting other factors may also play a role.

Correlation implies association, not causation. This relationship might be driven by reverse causality or omitted variables like education. Cross-sectional data cannot establish a definitive causal link.