---
title: "Assignment 2"
author: "Aleksandar Maric, Akando Räderer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(ggplot2)
```
# Assignment 2

## 1. Data preparation

### 1. Read-in, variable construction

At first we read-in the `.csv` files with `read.csv`. After a look at both datasets, we observe that `data_cleaned` is ready to go. However in `data_surnames` there is one row for Sri Lanka, which we will filter out.

```{r}
data_cleaned <- read.csv("data/ps2_cantons_cleaned.csv", sep = ";")
data_surnames <- read.csv("data/ps2_cantons_surnames.csv")

data_cleaned
data_surnames

data_surnames <- data_surnames |> filter(canton != "Sri Lanka") |> arrange(canton) #arranged it for better readability
```

Then we need to create the new variable `surname` and drop the existing variables for `share_mueller`and `share_maier`. 

```{r}
data_surnames <- data_surnames |> mutate(surname=share_mueller+share_maier) |> 
  select(-share_mueller, -share_maier)
```


### 2. Merging datasets

Before joining, we align the canton name that differs between the two datasets: `data_cleaned` uses `"St. Gallen"` while `data_surnames` uses `"Sankt Gallen"`.

```{r}

data_surnames <- data_surnames |>
  mutate(canton = ifelse(canton == "Sankt Gallen", "St. Gallen", canton))


data_joint <- data_cleaned |>
  left_join(data_surnames, by = "canton")
```


### 3. Data checks

We perform data checks: no missing `surname` values after the join, exactly 26 unique cantons, and only the required variables in the final dataset.

```{r}
sum(is.na(data_joint$surname))

n_distinct(data_joint$canton)

data_joint <- data_joint |> select(canton, gdp, population, immigration, surname)

head(data_joint)
```

There are no missing surnames and exactly 26 cantons are present in the dataset $\checkmark$.

## 2. Estimation of an univariate model (Effect of surname on gdp)

### 1. Regression equation

The regression equation for this specific model is:

$$\text{gdp}_i = \beta_0 + \beta_1 \cdot \text{surname}_i + u_i$$
where $\text{gdp}_i$ is the total GDP of canton $i$ in million CHF, $\text{surname}_i$ is the share of the cantonal population with the surname Mueller or Maier in percent, $\beta_0$ is the intercept, $\beta_1$ is the slope coeff., and $u_i$ is the error term.


### 2. Estimation of model coefficients

```{r}
model_surname <- lm(gdp ~ surname, data = data_joint)
print(model_surname$coefficients)
```
**Intercept ($\hat{\beta}_0$):** The estimated intercept is 10077.28. 

$\rightarrow$ This represents the predicted GDP (in million CHF) of a canton where the share of people named Mueller or Maier is zero. Since a zero surname share is unrealistic, the intercept has limited economic interpretation on its own.

**Slope ($\hat{\beta}_1$):** The estimated slope on `surname` is 449549.89 

$\rightarrow$ This means that a one unit increase in the share of people with the surname Mueller or Maier is associated with an increase in GDP of $\approx 450$ million CHF. Note that the coefficient appears large because a "one unit increase" in `surname` would mean going from 0% to 100% of the population being named Mueller or Maier.


### 3. Prediction of `gdp` based on average shares of surnames

```{r}
avg_surname <- mean(data_joint$surname)
avg_surname

predicted_gdp <- as.numeric(model_surname$coefficients[1] + model_surname$coefficients[2] * avg_surname)
predicted_gdp

# AI proposed using predict(), which gives the same result
```

At the average surname share of 0.041, the model predicts a GDP of 28'656 million CHF — roughly 18'579 million CHF higher than what the model would predict for a canton with no Mueller/Maier residents $\hat{\beta}_0$ = 10'077 million CHF).

## Exercise 3: Comparison of two univariate models

### 1. Regression

```{r}
model_immigration <- lm(gdp ~ log(immigration), data = data_joint)
print(model_immigration$coefficients)
```

### 2. $R^2$: Predictive power

```{r}
gdp_pred_surname <- model_surname$coefficients[1] + model_surname$coefficients[2] * data_joint$surname
SSR_surname <- sum((data_joint$gdp - gdp_pred_surname)^2)
SST <- sum((mean(data_joint$gdp) - data_joint$gdp)^2)
SSE <- SST - SSR_surname # for completeness
rsquared_surname <- 1 - SSR_surname / SST #alternatively SSE / SST

gdp_pred_immigration <- model_immigration$coefficients[1] + model_immigration$coefficients[2] * log(data_joint$immigration)
SSR_immigration <- sum((data_joint$gdp - gdp_pred_immigration)^2)
SSE_immigration = SST - SSR_immigration # for completeness
rsquared_immigration <- 1 - SSR_immigration / SST #atlernatively SSE_immigration / SST

print(rsquared_surname)
print(rsquared_immigration)
```

$R^2$ measures the share of total variation in cantonal GDP that is explained by the predictor. An $R^2$ of 1 means the predictor perfectly explains all variation in GDP, while an $R^2$ of 0 means it explains none.

The surname model explains 64.27% of the variation in GDP, while the log immigration model explains approximately 60.52%. The predictive power of the surname model is therefore higher.

### 3. Discussion on biasedness

For the OLS estimator to be unbiased, the zero conditional mean assumption (SLR.4) must hold: $E[u \mid X] = 0$. This requires that the predictor is uncorrelated with the error term, i.e., that no important variables are omitted that simultaneously affect GDP and the predictor.

**Surname model:** It is very unlikely that the share of people named Mueller or Maier satisfies SLR.4. Cantons with a higher surname share tend to be German-speaking cantons in central Switzerland, which also tend to differ systematically from French- and Italian-speaking cantons in terms of industrial structure, urban density, and historical development, which affect GDP independently. This is a classic case of omitted variable bias: the surname variable is picking up the effect of other canton characteristics, so $\hat{\beta}_1$ is likely biased.

**Log immigration model:** Immigration is also unlikely to satisfy SLR.4 in a univariate model. Cantons with higher immigration tend to be richer, more urban cantons (e.g., Zurich, Geneva), and this relationship is partly driven by reverse causality (higher GDP attracts immigrants) and by omitted variables such as population size, industry mix, and labor market conditions. Thus $\hat{\beta}_1$ in the immigration model is also likely biased.

In summary, neither coefficient can be given a causal interpretation without further controls.

### 4. Relationship between unbiasedness and the predictive power of a model

Unbiasedness and the predictive power measured by $R^2$ are two separate concepts that are not necessarily related. A model can fit the data very well (high $R^2$) while its coefficients are biased due to omitted variables, and an unbiased estimator can still have a low $R^2$ if the predictor only weakly explains the outcome. In short, $R^2$ tells us how well the model predicts, not whether we can trust the coefficients causally.









