---
title: "Assignment 2"
author: "Aleksandar Maric, Akando Räderer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(ggplot2)
```
# Assignment 2

## 1. Data preparation

### 1. Read-in, variable construction

At first we read-in the `.csv` files with `read.csv`. After a look at both datasets, we observe that `data_cleaned` is ready to go. However in `data_surnames` there is one row for Sri Lanka, which we will filter out.

```{r}
data_cleaned <- read.csv("ps2_cantons_cleaned.csv", sep = ";")
data_surnames <- read.csv("ps2_cantons_surnames.csv")

data_cleaned
data_surnames

data_surnames <- data_surnames |> filter(canton != "Sri Lanka") |> arrange(canton) #arranged it for better readability
```

Then we need to create the new variable `surname` and drop the existing variables for `share_mueller`and `share_maier`. 

```{r}
data_surnames <- data_surnames |> mutate(surname=share_mueller+share_maier) |> 
  select(-share_mueller, -share_maier)
```


### 2. Merging datasets

```{r}
setdiff(data_cleaned$canton, data_surnames$canton)
```

Before joining, we align the canton name that differs between the two datasets: `data_cleaned` uses `"St. Gallen"` while `data_surnames` uses `"Sankt Gallen"`.

```{r}
data_surnames$canton[data_surnames$canton == "Sankt Gallen"] <- "St. Gallen"

data_joint <- data_cleaned |>
  left_join(data_surnames, by = "canton")
```


### 3. Data checks

We perform data checks: no missing `surname` values after the join, exactly 26 unique cantons, and only the required variables in the final dataset.

```{r}
sum(is.na(data_joint$surname))

n_distinct(data_joint$canton)

data_joint <- data_joint |> select(canton, gdp, population, immigration, surname)

head(data_joint)
```

$\checkmark$

## 2. Estimation of an univariate model (Effect of surname on gdp)

### 1. Regression equation

The regression equation for this specific model is:

$$\text{gdp}_i = \beta_0 + \beta_1 \cdot \text{surname}_i + u_i$$
where $\text{gdp}_i$ is the total GDP of canton $i$ in million CHF, $\text{surname}_i$ is the share of the cantonal population with the surname Mueller or Maier in percent, $\beta_0$ is the intercept, $\beta_1$ is the slope coeff., and $u_i$ is the error term.


### 2. Estimation of model coefficients

```{r}
model_surname <- lm(gdp ~ surname, data = data_joint)
print(model_surname$coefficients)
```
**Intercept ($\hat{\beta}_0$):** The estimated intercept is 10077.28. 

$\rightarrow$ This represents the predicted GDP (in million CHF) of a canton where the share of people named Mueller or Maier is zero. Since a zero surname share is unrealistic, the intercept has limited economic interpretation on its own.

**Slope ($\hat{\beta}_1$):** The estimated slope on `surname` is 449549.89 

$\rightarrow$ This means that a one unit increase in the share of people with the surname Mueller or Maier is associated with an estimated increase in GDP of 449'549.89 million CHF. Note that the coefficient appears large because a "one unit increase" in `surname` would mean going from 0% to 100% of the population being named Mueller or Maier.


### 3. Prediction of `gdp` based on average shares of surnames

```{r}
# Average surname share across all cantons in the sample
avg_surname <- weighted.mean(data_joint$surname, w = data_joint$population)
avg_surname

# Predict GDP at the average surname share
predicted_gdp <- as.numeric(model_surname$coefficients[1] + model_surname$coefficients[2] * avg_surname)
predicted_gdp

# avg_surname <- mean(data_joint$surname)
# avg_surname

# Due to unclear instructions about the average, we decided to use the weighted average as this reflects the effect of the relative population of each canton on GDP more accurately. The absolute difference in avg_surname is approx. 0.05%.
```

At the average cantonal surname share of `r round(avg_surname, 4)` %, the model predicts a GDP of approximately 51'512.57 million CHF.


## Comparison of two univariate models

### 1. Regression

```{r}
model_immigration <- lm(gdp ~ log(immigration), data = data_joint)
print(model_immigration$coefficients)
```

### 2. $R^2$: Predictive power

```{r}
gdp_pred_surname <- model_surname$coefficients[1] + model_surname$coefficients[2] * data_joint$surname
SSR_surname <- sum((data_joint$gdp - gdp_pred_surname)^2)
SST <- sum((mean(data_joint$gdp) - data_joint$gdp)^2)
SSE <- SST - SSR_surname # for completeness
rsquared_surname <- 1 - SSR_surname / SST #alternatively SSE / SST

gdp_pred_immigration <- model_immigration$coefficients[1] + model_immigration$coefficients[2] * log(data_joint$immigration)
SSR_immigration <- sum((data_joint$gdp - gdp_pred_immigration)^2)
SSE_immigration = SST - SSR_immigration # for completeness
rsquared_immigration <- 1 - SSR_immigration / SST #atlernatively SSE_immigration / SST

print(rsquared_surname)
print(rsquared_immigration)
```

$R^2$ measures the share of total variation in cantonal GDP that is explained by the predictor. An $R^2$ of 1 means the predictor perfectly explains all variation in GDP, while an $R^2$ of 0 means it explains none.

The surname model explains 64.27% of the variation in GDP, while the log immigration model explains approximately 60.52%. The predictive power of the surname model is therefore higher.

### 3. Discussion on biasedness

For the OLS estimator to be unbiased, the zero conditional mean assumption (SLR.4) must hold: $E[u \mid X] = 0$. This requires that the predictor is uncorrelated with the error term, i.e., that no important variables are omitted that simultaneously affect GDP and the predictor.

**Surname model:** It is very unlikely that the share of people named Mueller or Maier satisfies SLR.4. Cantons with a higher surname share tend to be German-speaking cantons in central Switzerland, which also tend to differ systematically from French- and Italian-speaking cantons in terms of industrial structure, urban density, and historical development, which affect GDP independently. This is a classic case of omitted variable bias: the surname variable is picking up the effect of other canton characteristics, so $\hat{\beta}_1$ is likely biased.

**Log immigration model:** Immigration is also unlikely to satisfy SLR.4 in a univariate model. Cantons with higher immigration tend to be richer, more urban cantons (e.g., Zurich, Geneva), and this relationship is partly driven by reverse causality (higher GDP attracts immigrants) and by omitted variables such as population size, industry mix, and labor market conditions. Thus $\hat{\beta}_1$ in the immigration model is also likely biased.

In summary, neither coefficient can be given a causal interpretation without further controls.

### 4. Relationship between unbiasedness and the predictive power of a model

Unbiasedness and the predictive power measured by $R^2$ are two separate concepts that are not necessarily related. A model can fit the data very well (high $R^2$) while its coefficients are biased due to omitted variables, and an unbiased estimator can still have a low $R^2$ if the predictor only weakly explains the outcome. In short, $R^2$ tells us how well the model predicts, not whether we can trust the coefficients causally.

## 4. Univariate model without a constant

### 1. Formula

$$\text{gdp}_i = \beta_1 \cdot \text{surname}_i + u_i$$

### 2. Estimation of the model

```{r}
model_no_constant <- lm(gdp ~ surname - 1, data = data_joint)
summary(model_no_constant)
```
The estimated coefficient on `surname` is 532'247. (*From the function documentation: A formula has an implied intercept term. To remove this use either y ~ x - 1 or y ~ 0 + x.*)

### 3. Explanation for the change

When the intercept is removed, OLS is forced to fit the regression line through the origin (0, 0). Since the data does not actually pass through the origin, the slope coefficient absorbs part of what the intercept would have captured, causing it to change.


## 5. Own OLS estimator

### 1. Function

```{r}
my_ols <- function(y, x) {
    # means
    x_bar <- mean(x)
    y_bar <- mean(y)
    
    # apply formula for both coefficients
    beta_1 <- sum((y - y_bar) * (x - x_bar)) / sum((x - x_bar)^2) # Cov(y,x) / Var(x)

    beta_0 <- y_bar - beta_1 * x_bar
    
    return(c(Intercept = beta_0, Slope = beta_1))
}
```

### 2. Verfication against `lm()`

```{r}
my_ols(y = data_joint$gdp, x = data_joint$surname)

# lm() for comparison
coef(model_surname)
```

Both sets of coefficients match, confirming that the function is correctly implemented.

### 3. `if` condition for case without a constant

```{r}
my_if_ols <- function(y, x, add_constant = TRUE) {
  
  if (add_constant) {
    
    # means
    x_bar <- mean(x)
    y_bar <- mean(y)
    
    # apply formula for both coefficients
    beta_1 <- sum((y - y_bar) * (x - x_bar)) / sum((x - x_bar)^2) # Cov(y,x) / Var(x)

    beta_0 <- y_bar - beta_1 * x_bar
    
    return(c(Intercept = beta_0, Slope = beta_1))
    
  } else {

    # look at derivation below 
    beta_1 <- sum(x * y) / sum(x^2)
    
    return(c(Slope = beta_1))
  }
}
```

Now to verify it, we compare it against the results from subexample 4:

```{r}
my_if_ols(y = data_joint$gdp, x = data_joint$surname, add_constant = FALSE)
coef(model_no_constant)
```

Both slopes again match, confirming that the `add_constant = FALSE` branch works correctly.



#### Appendix
If there is no intercept the FOC for $\hat{\beta}_1$ becomes: 
$$
\begin{aligned}
\frac{\partial}{\partial \beta_1} \sum_{i=1}^n (y_i - \beta_1 x_i)^2 &\overset{!}{=} 0 \\
\sum_{i=1}^{n} 2(y_i - \beta_1 x_i) \cdot (-x_i) &= 0 \\
\sum_{i=1}^{n} x_i(y_i - \beta_1 x_i) &= 0 \\
\sum_{i=1}^{n} x_i y_i - \beta_1 \sum_{i=1}^{n} x_i^2 &= 0 \\
\implies \hat{\beta}_1 &= \frac{\sum_{i=1}^{n} x_i y_i}{\sum_{i=1}^{n} x_i^2}
\end{aligned}
$$
