---
title: "Assignment 2"
author: "Aleksandar Maric, Akando Räderer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(ggplot2)
```
# Assignment 2

## 1. Data preparation

### 1. Read-in, variable construction

At first we read-in the `.csv` files with `read.csv`. After a look at both datasets, we observe that `data_cleaned` is ready to go. However in `data_surnames` there is one row for Sri Lanka, which we will filter out.

```{r}
data_cleaned <- read.csv("data/ps2_cantons_cleaned.csv", sep = ";")
data_surnames <- read.csv("data/ps2_cantons_surnames.csv")

data_cleaned
data_surnames

data_surnames <- data_surnames |> filter(canton != "Sri Lanka") |> arrange(canton) #arranged it for better readability
```

Then we need to create the new variable `surname` and drop the existing variables for `share_mueller`and `share_maier`. 

```{r}
data_surnames <- data_surnames |> mutate(surname=share_mueller+share_maier) |> 
  select(-share_mueller, -share_maier)
```


### 2. Merging datasets

Before joining, we align the canton name that differs between the two datasets: `data_cleaned` uses `"St. Gallen"` while `data_surnames` uses `"Sankt Gallen"`.

```{r}

data_surnames <- data_surnames |>
  mutate(canton = ifelse(canton == "Sankt Gallen", "St. Gallen", canton))


data_joint <- data_cleaned |>
  left_join(data_surnames, by = "canton")
```


### 3. Data checks

We perform data checks: no missing `surname` values after the join, exactly 26 unique cantons, and only the required variables in the final dataset.

```{r}
sum(is.na(data_joint$surname))

n_distinct(data_joint$canton)

data_joint <- data_joint |> select(canton, gdp, population, immigration, surname)

head(data_joint)
```

There are no missing surnames and exactly 26 cantons are present in the dataset $\checkmark$.

## 2. Estimation of an univariate model (Effect of gdp on surname)

### 1. 

The regression equation for this specific model is:

$$gdp_i = \beta_0 + \beta_1 \cdot surname_i + u_i$$
where $gdp_i$ is the total GDP of canton $i$ in million CHF, $surname_i$ is the share of the cantonal population with the surname Mueller or Maier in percent, $\beta_0$ is the intercept, $\beta_1$ is the slope coeff., and $u_i$ is the error term.


### 2. Estimation of model coefficients

```{r}
model_surname <- lm(gdp ~ surname, data = data_joint)
print(model_surname$coefficients)
```
**Intercept ($\hat{\beta}_0$):** The estimated intercept is 10077.28. 

$\rightarrow$ This represents the predicted GDP (in million CHF) of a canton where the share of people named Mueller or Maier is zero. Since a zero surname share is unrealistic, the intercept has limited economic interpretation on its own.

**Slope ($\hat{\beta}_1$):** The estimated slope on `surname` is 449549.89 

$\rightarrow$ This means that a one unit increase in the share of people with the surname Mueller or Maier is associated with an increase in GDP of $\approx 450$ million CHF. Note that the coefficient appears large because a "one unit increase" in `surname` would mean going from 0% to 100% of the population being named Mueller or Maier — an impossible scenario — so the number should be interpreted at realistic, much smaller increments (explained by Claude.ai).


### 3. Prediction of `gdp` based on average shares of surnames

```{r}
avg_surname <- mean(data_joint$surname)
avg_surname

# Predict GDP at the average surname share
predicted_gdp <- as.numeric(model_surname$coefficients[1] + model_surname$coefficients[2] * avg_surname)
predicted_gdp

# AI proposed using predict(), which gives the same result
```

At the average cantonal surname share of 0.041%, the model predicts a GDP of approximately 28'656 million CHF









