---
title: "Data Analytics II: Empirical Research Methods - Exercise 1"
author:
- Aleksandar Maric
- Akando Räderer
- Daniel Margolin
date: "28.02.2026"
output:
  pdf_document:
    latex_engine: xelatex
---

### Exercise 1a

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
options(tinytex.verbose = TRUE)
```


```{r}
data <- read.csv("Ex1.csv")
summary(data)
```

Renaming the misspelled columns with

```{r}
data <- data |> rename(color=c0lor)
data <- data |> rename(price=prise)
```

Removing an unnecessairy information from the data set -> color

```{r}
data <- data |> select(-color)
```

Now we check for missing values 

```{r}
skimr::skim(data)
```
and wee see, that in the column "clarity" there are almost 5k missing entries. Removing them finally gives us a clean dataset.

```{r}
data <- data |> filter(!is.na(clarity))
```

Now we compute the averages and the standard deviations:

```{r}
mean(data$carat)
sd(data$carat)
mean(data$price)
sd(data$price)
```
And finally produce a density plot to visualize the distribution of price per carat, while grouped by the cut. 

```{r}
data$cut <- factor(data$cut, levels = c("Fair", "Good", "Very Good", "Premium", "Ideal"), ordered = TRUE)
```

With `ggplot` we can plot the graphs. With `facet_wrap()` we can create plots for each category we just split the data into. The scale is also changed to logarithmic for better readability.

```{r}
ggplot(data = data, aes(x=price/carat))+
  geom_density()+
  facet_wrap(~(cut), scale = "free")+
  scale_x_log10()+
  ylim(0, 11) + 
  ggtitle("Distribution of Diamond Price per Carat: grouped by cut.") + 
  xlab("Diamond Price per Carat U$ - LOG 10 Scale") + 
  ylab("Density (scaled)") + 
  theme_minimal()
```


From the chart, we can observe that the price per carat increases with cut quality. However, this result does not imply that cut quality increases price / carat causally. Further analysis would be needed to draw conclusions in this regard. You will learn the tools to conduct such analyses during the rest of the semester.





